Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}

### 2017-10-13 conclusion.tex
Typos

We have presented qNML, a new model selection criterion for learning
structures of Bayesian networks.  While being competitive in
predictive terms, it often yields significantly simpler models than
other common model selection criteria other than BIC that has a very strong
bias for simplicity. The computational cost of qNML
equals the cost of the current state-of-the-art criteria. The
criterion also gives equal scores for models that encode same
independence hypotheses about the joint probability distribution.
qNML also coincides with NML criterion for many models.
In our experiments, the qNML criterion appears as a safe choice for a
model selection criterion that balances
parsimony, predictive capability and the ability to quickly converge to the
generating model.

>>>>

We have presented qNML, a new model selection criterion for learning
structures of Bayesian networks.  While being competitive in
predictive terms, it often yields significantly simpler models than
other common model selection criteria other than BIC that has a very strong
bias for simplicity. The computational cost of qNML
equals the cost of the current state-of-the-art criteria. The
criterion also gives equal scores for models that encode the same
independence hypotheses about the joint probability distribution.
qNML also coincides with the NML criterion for many models.
In our experiments, the qNML criterion appears as a safe choice for a
model selection criterion that balances
parsimony, predictive capability and the ability to quickly converge to the
generating model.

### 2017-10-13 supp_mat_B.tex
Typos

\subsection{fNML is Regular}
In this section, we will show that fNML score is regular. Recall first the definition of fNML local score
\begin{equation}
Q^{fnml}_N(X  \vert \ U) = \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \sum_{j = 1}^{r_U} reg(N_j,r_X),
\end{equation}where $r_X$ is the number of categories for $X$, $r_U$ denotes the number of possible configurations of variables in $U$, and $N_j$ is the number of times the $j$:th configuration is observed in our samples $u_N$. Note that $reg(r_X, N_j) = 0$ for any configuration not observed in our sample. 
fNML criterion differs from qNML only by how the penalty term is defined. We can follow a highly similar strategy in order to prove the regularity for fNML. To this end, we need the following Lemma:
\begin{lemma}\label{regineq}
Let $N = N_1 + N_2$ and $r \in\N, r > 2$. Now,
$$
reg(N,r) \leq reg(N_1,r) + reg(N_2,r).
$$
\end{lemma}
\begin{proof}
We start with the definition of $C(N_1,r) = \exp(reg(N_1,r))$,
\begin{align*}
C(N_1,r)C(N_2,r) &= \sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N_1} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N_2} ))\\ & \geq
\sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N} )) \\ &=
\sum_{x_{N}}P(x_{N} \vert \hat{\theta}(x_{N} )) \\ & = C(N,r).
\end{align*}Taking the logarithm on the both sides yields our claim. On the second row, we used the definition of maximum likelihood parameters: the probability of data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N_1})$, probability can only go down or stay the same. On the third row, we used the i.i.d, assumption, which allows us to take the single sum over data vectors of length $N = N_1 + N_2$.     
\end{proof}
Now we can proceed to the actual proof.
\begin{theorem}
fNML score is regular.
\end{theorem}
\begin{proof}
Using the entropy assumption and Lemma \ref{C2}, we can ignore the maximized likelihood terms and study the penalty terms. We want show that
\begin{equation}\label{fnmlPenalty}
\sum_{j = 1}^{r_{Y'}} reg(N'_j,r_X) \leq \sum_{l = 1}^{r_{Y}} reg(N_l,r_X),
\end{equation}where we used $r_{Y'}$ and $r_Y$ to denote the numbers of possible configurations for variables in $Y'$ and $Y$, respectively. Also, $N_j$ refers to the number times the $j$:th configuration of variables $Y$ is observed in our sample, and $N'_l$ denotes the same for $Y'$.

>>>>

\subsection{fNML is Regular}
In this section, we will show that fNML score is regular. Recall first the definition of fNML local score
\begin{equation}
Q^{fnml}_N(X  \vert \ U) = \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \sum_{j = 1}^{r_U} reg(N_j,r_X),
\end{equation}where $r_X$ is the number of categories for $X$, $r_U$ denotes the number of possible configurations of variables in $U$, and $N_j$ is the number of times the $j$:th configuration is observed in our samples $u_N$. Note that $reg(r_X, N_j) = 0$ for any configuration not observed in our sample. 
fNML criterion differs from qNML only by how the penalty term is defined. We can follow a highly similar strategy in order to prove the regularity for fNML. To this end, we need the following Lemma:
\begin{lemma}\label{regineq}
Let $N = N_1 + N_2$ and $r \in\N, r > 2$. Now,
$$
reg(N,r) \leq reg(N_1,r) + reg(N_2,r).
$$
\end{lemma}
\begin{proof}
We start with the definition of $C(N_1,r) = \exp(reg(N_1,r))$,
\begin{align*}
C(N_1,r)C(N_2,r) &= \sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N_1} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N_2} ))\\ & \geq
\sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N} )) \\ &=
\sum_{x_{N}}P(x_{N} \vert \hat{\theta}(x_{N} )) \\ & = C(N,r).
\end{align*}Taking the logarithm on both sides yields our claim. On the second row, we used the definition of maximum likelihood parameters: the probability of the data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N_1})$, the probability can only go down or stay the same. On the third row, we used the i.i.d., assumption, which allows us to take the single sum over data vectors of length $N = N_1 + N_2$.
\end{proof}
Now we can proceed to the actual proof.
\begin{theorem}
fNML score is regular.
\end{theorem}
\begin{proof}
Using the entropy assumption and Lemma \ref{C2}, we can ignore the maximized likelihood terms and study the penalty terms. We want show that
\begin{equation}\label{fnmlPenalty}
\sum_{j = 1}^{r_{Y'}} reg(N'_j,r_X) \leq \sum_{l = 1}^{r_{Y}} reg(N_l,r_X),
\end{equation}where we used $r_{Y'}$ and $r_Y$ to denote the numbers of possible configurations for variables in $Y'$ and $Y$, respectively. Also, $N_j$ refers to the number times the $j$:th configuration of variables $Y$ is observed in our sample, and $N'_l$ denotes the same for $Y'$.
